---
title: "quadratic equation"
format: html
author: Veronika Post
editor: visual
---

# Quadratic Equation

*High school lesson*

Let's see how we can apply the quadratic equation and graphically demonstrate that the solution is correct.

We have an equation:

$$
f(x) = ax^2 + bx + c = 0
$$

Let's define variables for this equation:

```{r}
a <- 1
b <- 3
c = 2
```

To find real solutions we can use the following equation:

$$
x = \frac{-b \pm \sqrt{b^2 -4ac}}{2a}
$$

Let's use code to find values of $x$ programmatically:

```{r}
# We have to account for the plus-minus sign in our solution

# First, let's calculate the value that is under the square sign - 
# it is called "discriminant".

discr <- b^2 - 4*a*c

# If discriminant is negative - we don't have any real solutions as there are 
# no real numbers that would result in a negative square root
if (discr < 0) {
  cat("There are no real solutions.")
  
# If discriminant is positive we can calculate x1 and x2
}else {
  x_1 <- (-b + sqrt(discr)) / 2*a
  x_2 <- (-b - sqrt(discr)) / 2*a
  cat("The real solutions are x1:", x_1, "and x2:", x_2)
}
```

We can check whether the values that we found are correct by plotting a graph of our function. On the plot below we can see two points where the curve of our function crosses the horizontal line where $y = 0$. These points are $(-1, 0)$ and $(2, 0)$.

```{r}

library(ggplot2)

# We can use the 'ggplot2` library. To do so we can
# create a dataframe of the x and y values.

# To plot this function for values of x from -3 to 3 we have 
# to calculate y values for each corresponding x using the formula above:
xs <- seq(-3, 3, length.out = 100)
ys <- a * xs^2 + b * xs + c

data <- data.frame(x = xs, y = ys)

# Note: I'm using geom_smooth to plot a smooth line instead of points 
# - there might be a better way to do so.
ggplot(data, aes(x, y)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "transparent") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  geom_point(aes(x = x_1, y = 0), color = "blue", size = 3) + 
  geom_point(aes(x = x_2, y = 0), color = "blue", size = 3) +
  scale_x_continuous(breaks = seq(-3, 3, 1)) +
  labs(title = expression(paste("Graph of ", ax^2 + bx + c)),
       x = "x", y = "y")
```
